---
title: "Risk Analytics - Practical 1"
subtitle : "Winter semester 2024-2025, HEC, UNIL"
author: "Robin Michel, Faber Bickerstaffe, Antoine Magnin, Anastasia Pushkarev and Victorien Rodondi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Load libraries
library(readr)
library(tseries)
library(nortest)
library(MASS)
library(fpp2)
library(fGarch)
library(lmtest)
library(ggplot2)
library(tidyr)
library(vars)

# Load data using a relative path
Crypto_data <- read_csv("Data/Crypto_data.csv")
```

# Part 1: Financial Returns and Normality

### a) and b) Assessing Stationarity of Bitcoin Prices and Negative Log Returns

In this section, we assess the stationarity of the raw Bitcoin prices and their negative log returns (see Figure 1), as stationarity is crucial for time series modeling.

We first plotted the Bitcoin prices (see Figure 2) to visually inspect for trends or patterns. The plot showed a clear trend, suggesting non-stationarity. To confirm this, we applied the Augmented Dickey-Fuller (ADF) test (see Table 1), which resulted in a p-value of 0.3885, indicating that the raw Bitcoin prices are non-stationary.

To address this, we computed the negative log returns, a transformation commonly used in financial time series analysis to obtain a stationary series. Visual inspection of the negative log returns (see Figure 3) suggested stationarity. This was further confirmed by the ADF test (see Table 2), which gave a p-value of 0.01, leading us to reject the null hypothesis of non-stationarity and confirming that the negative log returns are stationary.

### c) Assessing the Normality of Negative Log Returns

To evaluate whether the negative log returns of Bitcoin follow a normal distribution, we first examined the data visually using a histogram (see Figure 4) and a QQ-plot (see Figure 5). The histogram of the negative log returns suggests that the data may be approximately normally distributed, though some deviations from normality could exist. Additionally, the QQ-plot shows that the returns are normally distributed for non-extreme values, but there are noticeable deviations in the tails, indicating that the negative log returns may not follow a perfect normal distribution.

To formally test for normality, we applied the Anderson-Darling test (see Table 3), which gave a p-value less than 0.05. As a result, we reject the null hypothesis (H0) that the data is normally distributed. This confirms that, despite appearing somewhat normal in the central part of the distribution, the negative log returns are not normally distributed, especially due to the extreme values.

### d) and e) Fitting a t-Distribution and Comparing Tails

Since the negative log returns deviate from normality, particularly in the extremes, we fit a t-distribution to the scaled data to check if it better captures these extreme values. A QQ-plot was generated to compare the negative log returns with the theoretical t-distribution (see Figure 6), which showed that the data fits the t-distribution quite well, including in the tails. For comparison, we also generated a QQ-plot for the normal distribution (see Figure 5), which demonstrated a poorer fit, particularly for extreme values. This suggests that the t-distribution, with its ability to model heavy tails, is a more appropriate fit for the data (see Figure 8).

Next, we compared the density plots of the normal and t-distributions. As expected, the t-distribution exhibited heavier tails than the normal distribution, meaning we should expect more extreme, unexpected events in a t-distribution (see Figure 9).

Since the Bitcoin data follows the t-distribution more closely, and the t-distribution has fatter tails than the normal distribution, this indicates that extreme values (large deviations from the mean) are more likely in the Bitcoin data than if it were normally distributed.

# Part 2: Financial time series, heteroscedasticity and the random walk hypothesis

## a) and b) Assessing Autocorrelation in Bitcoin Prices

Building on the previous stationarity analysis, we further explored the autocorrelation of the raw Bitcoin prices and the negative log returns. In part (a), we plotted the ACF (Autocorrelation Function) for both series (see Figures 9 and 10). As expected, the raw Bitcoin prices showed strong and persistent autocorrelation, consistent with their non-stationary nature. In contrast, the negative log returns displayed only minor, short-lived autocorrelation, aligning with their stationary behavior.

In part (b), we applied the Ljung-Box test (see Tables 4 and 5) to formally assess autocorrelation. The test confirmed significant autocorrelation in both series (p-value < 0.05). However, the negative log returns exhibited much weaker and short-term autocorrelation compared to the raw prices. This suggests that, while the log returns are stationary, they still contain some mild autocorrelation, typical of financial time series.

In summary, the raw series has strong autocorrelation and is non-stationary, whereas the negative log returns are stationary with only moderate, short-lived autocorrelation, making them better suited for modeling.


## c) ARIMA Modeling

In this section, we proposed an ARIMA model for the negative log returns based on the analysis of the ACF and PACF plots (see Figures 10 and 11). The ACF plot showed a spike at lag 2, suggesting a moving average component of q = 2. Similarly, the PACF plot indicated a significant spike at lag 2, suggesting an autoregressive component of p = 2. Since the series is already stationary, we set d = 0. Based on this, we fit an ARIMA(2, 0, 2) model to the negative log returns. To confirm our model selection, we used the auto.arima() function, which automatically identifies the best ARIMA model. The function also suggested an ARIMA(2, 0, 2) model, reinforcing our choice.

Next, we assessed the residuals of the ARIMA model to evaluate its adequacy. The ACF of the residuals showed no significant autocorrelation (see Figure 12), indicating that the model successfully captured the underlying patterns in the data. The Ljung-Box test further confirmed this (see Table 6), as the p-value was greater than 0.05, suggesting no significant autocorrelation remaining in the residuals. However, the QQ-plot (see Figure 13) and Shapiro-Wilk test (see Table 7) indicated that the residuals do not follow a normal distribution (p-value < 0.05). Additionally, the plot of residuals over time (see Figure 14) showed signs of volatility clustering, implying that the variance is not constant, which indicates the presence of heteroscedasticity. This suggests that while the ARIMA model adequately captures the autocorrelation in the data, it fails to account for the changing variance and heavy tails in the residuals.

In conclusion, the ARIMA(2, 0, 2) model effectively captures the serial dependencies in the negative log returns but may not fully address the volatility present in the data. This indicates that a GARCH model, which accounts for time-varying volatility, could be a better fit for modeling the residuals.

## d) GARCH Modeling

In this section, we fitted GARCH(1,1) models to the negative log returns using both a normal distribution and a standardized t-distribution for the residuals, aiming to capture the time-varying volatility (heteroscedasticity) of the residuals observed in the data.

Both models effectively captured most of the autocorrelation (see Figures 15 and 16), with only a small spike remaining at lag 1. The Ljung-Box test indicated some residual autocorrelation (p-value < 0.05) for both models (see Tables 8 and 9), suggesting minor dependencies left unmodeled. However, the QQ-plots revealed a notable difference (see Figures 17 and 18). The normal GARCH model showed clear deviations in the tails, indicating that the residuals deviate from a normal distribution and that the model did not handle extreme values well. The Shapiro-Wilk test (see Table 10) reject normality for the residuals of the normal GARCH model (p-value < 0.05), further confirming that the residuals do not align with a normal distribution and that the model is inadequate for capturing heavy tails. In contrast, the t-GARCH model's QQ-plot aligned better with the t-distribution quantiles, including in the tails, indicating that it better captured the heavy tails of the data. This confirms that the t-GARCH model more effectively handles extreme values compared to the normal GARCH model.

Despite the differences in tail behavior, the AIC/BIC values for both models were very close (AIC: -10.60 for normal vs. -10.75 for t-distribution), making it difficult to choose one model based on these criteria alone. However, the better fit of the t-GARCH model to the tail behavior suggests it is more appropriate for the data, as it captures the heavy tails more effectively than the normal GARCH model.

## e) Two-Step ARIMA-GARCH Modeling

To address the potential issue of residual serial correlation when fitting a GARCH model directly on the negative log returns, we applied a two-step approach as suggested. First, we fitted an ARIMA(2,0,2) model to the negative log returns to capture any autocorrelation in the data. We then extracted the residuals from this ARIMA fit and applied a GARCH(1,1) model to these residuals.

The Box-Ljung test on the residuals of the final GARCH model (see Table 11) showed a non-significant result (p-value=0.9365), indicating that there was no significant autocorrelation left in the residuals. This confirms that the two-step ARIMA-GARCH approach effectively eliminated serial correlation, solving the issue of residual autocorrelation seen when fitting a GARCH model directly on the negative log returns, and leading to a better specified model.

## f) Compare the Three Models

We compared the three models: ARIMA(2, 0, 2), GARCH(1, 1) with a normal distribution, and GARCH(1, 1) with a t-distribution, to find the best fit for the Bitcoin negative log returns.

The ARIMA model effectively captured the autocorrelation but struggled with the changing variance (heteroscedasticity), as evidenced by volatility clustering in the residuals. This indicates a violation of the homoscedasticity assumption, since ARIMA cannot model time-varying variance. Additionally, the ARIMA model failed to account for the heavy tails in the data, as its residuals showed deviations from a normal distribution in the extremes.

In contrast, the GARCH models handled the heteroscedasticity issue well, as they are specifically designed to model time-varying variance. The normal GARCH model captured most of the autocorrelation but did not fit the heavy tails effectively, showing deviations in the extremes. The t-GARCH model, however, provided a better fit for the tail behavior, as indicated by the alignment with the t-distribution quantiles in the QQ-plot.

Overall, the t-GARCH model is the most suitable, effectively handling both the heteroscedasticity and the heavy tails typical in financial time series, unlike the ARIMA model and the normal GARCH model.

# Part 3: Dependence between time series 

## a) Negative Log Returns of Bitcoin and Ethereum Dependence

To assess whether the negative log returns of Bitcoin and Ethereum are dependent, we conducted a correlation test using the cor.test() function. The result of the test showed a p-value of 0.905 (see Table 12), which is much higher than the common significance level of 0.05. Therefore, we fail to reject the null hypothesis. This suggests that there is no significant correlation between the negative log returns of Bitcoin and Ethereum, and we can conclude that the series are likely independent based on the available data.

## b) Cross-Correlation Function (CCF) between Bitcoin and Ethereum

We also examine the potential dependency between the negative log returns of Bitcoin and Ethereum, we calculated the cross-correlation function (CCF). The CCF measures the correlation between two time series at different lags, helping us identify if changes in one series are systematically followed by changes in the other.

The CCF plot (see Figure 19) showed no significant correlation at lag 0, indicating that the two series are not contemporaneously correlated. However, there was a notable spike at lag -5, suggesting a statistically significant correlation where changes in Ethereum's log returns precede changes in Bitcoin's log returns by about 5 periods. This indicates some degree of dependency between the series, with Ethereum potentially influencing Bitcoin at certain points in time based on this CCF analysis.

## c) Granger Causality Test

Based on the Granger causality tests, we assessed whether the negative log returns of Bitcoin and Ethereum have predictive power for one another. Using a lag order of 6 (as determined by the VAR lag selection criteria in Table 13), the results indicate the following:

**Bitcoin predicting Ethereum:** The test yielded a very small p-value (0.001) (see Table 13), leading us to reject the null hypothesis that Bitcoin's past returns have no predictive power for Ethereum. This suggests that Bitcoin's returns can be used to forecast Ethereum's returns.

**Ethereum predicting Bitcoin:** The test yielded a large p-value (0.81) (see Table 14), so we fail to reject the null hypothesis. This indicates that Ethereum's past returns do not provide any predictive power for Bitcoin's future returns.

The Granger causality analysis shows that Bitcoin has predictive power for Ethereum, but not the other way around. This suggests a unidirectional relationship where Bitcoin influences Ethereum, but changes in Ethereum do not predict Bitcoin's returns.


## d) Interpretation of Granger Causality Results

Based on the Granger causality results, we can infer the following:

**If Bitcoin experiences a sudden drop:** We would expect Ethereum to also be negatively impacted shortly after. The Granger causality analysis showed that Bitcoin's price movements have predictive power over Ethereum's, meaning a sharp decline in Bitcoin prices is likely to lead to a downturn in Ethereum's prices in the near future.

**If Ethereum experiences a sudden drop:** There is no expectation that Bitcoin will be similarly affected. The lack of Granger causality from Ethereum to Bitcoin suggests that Ethereum's price changes do not significantly predict Bitcoin's movements. Therefore, Bitcoin may not experience a similar drop and could remain stable or be influenced by other unrelated market factors.

# Appendices

## Practical 1

### Figures

#### Figure 1: Bitcoin Prices and Negative Log Returns Over Time on Common Scale

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Standardize both series (subtract mean, divide by standard deviation)
bitcoin_prices_standardized <- scale(bitcoin_prices)
bitcoin_negative_log_returns_standardized <- scale(bitcoin_negative_log_returns)

# Create a data frame combining both series for plotting
bitcoin_data_standardized <- data.frame(
  Time = 1:length(bitcoin_prices_standardized),
  Bitcoin_Prices_Standardized = bitcoin_prices_standardized,
  Negative_Log_Returns_Standardized = c(NA, bitcoin_negative_log_returns_standardized)  # Add NA to align lengths
)

# Reshape the data for ggplot
bitcoin_data_long_standardized <- pivot_longer(bitcoin_data_standardized, 
                                               cols = c(Bitcoin_Prices_Standardized, Negative_Log_Returns_Standardized), 
                                               names_to = "Series", values_to = "Values")

# Plot the standardized series on a common scale, ensuring raw prices are drawn last (on top)
ggplot() +
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Negative_Log_Returns_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Bitcoin_Prices_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  labs(title = "Comparison of Standardized Bitcoin Prices and Negative Log Returns",
       x = "Time", y = "Standardized Values") +
  scale_color_manual(values = c("Bitcoin_Prices_Standardized" = "blue", 
                                "Negative_Log_Returns_Standardized" = "red"),
                     labels = c("Prices", "Log Returns")) +  # Shortened labels
  theme_minimal()
```


#### Figure 2: Bitcoin Prices Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Plot the Bitcoin prices to visually inspect stationarity
plot(bitcoin_prices, type = "l", main = "Bitcoin Prices", xlab = "Time", ylab = "Price")
```

#### Figure 3: Negative Log Returns of Bitcoin Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Plot the negative log returns
plot(bitcoin_negative_log_returns, type = "l", main = "Negative Log Returns of Bitcoin", xlab = "Time", ylab = "Negative Log Returns")
```


#### Figure 4: Histogram of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Draw a histogram of the negative log returns
hist(bitcoin_negative_log_returns, breaks = 30, main = "Histogram of Negative Log Returns", xlab = "Negative Log Returns", col = "lightblue", border = "black")
# --> seems normally distributed
```

#### Figure 5: QQ-Plot of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a QQ-plot to check normality
qqnorm(bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs. Normal Distribution")
qqline(bitcoin_negative_log_returns, col = "red") # --> normally distributed only for non-extreme values
```

#### Figure 6: QQ-Plot of Negative Log Returns with t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
### QQ-plot for t-distribution : seem to follow it quite well
# Generate a QQ-plot for the t-distribution fit
qqplot(qt(ppoints(length(bitcoin_negative_log_returns)), df = df_value), 
       bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs t-Distribution",
       xlab = "Theoretical Quantiles (t-distribution)", ylab = "Sample Quantiles")
# Add a 45-degree line to the QQ-plot
qqline(bitcoin_negative_log_returns, distribution = function(p) qt(p, df = df_value), col = "red")
```

#### Figure 7: Histogram of Bitcoin Negative Log Returns with Fitted t and Normal Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Extract the fitted parameters from t_fit
m <- t_fit$estimate["m"]      # Mean (location parameter)
s <- t_fit$estimate["s"]      # Scale (related to standard deviation)
df <- t_fit$estimate["df"]    # Degrees of freedom

# Calculate the mean and standard deviation for the normal distribution
mean_normal <- mean(bitcoin_negative_log_returns_scaled)
sd_normal <- sd(bitcoin_negative_log_returns_scaled)

# Plot the histogram of your data
hist(bitcoin_negative_log_returns_scaled, breaks = 30, probability = TRUE, 
     main = "Histogram of Bitcoin Neg. Log Ret. with Fitted t and Normal Distribution",
     xlab = "Negative Log Returns", col = "lightblue", border = "black")

# Overlay the density of the fitted t-distribution
x_vals <- seq(min(bitcoin_negative_log_returns_scaled), max(bitcoin_negative_log_returns_scaled), length.out = 1000)
t_density <- dt((x_vals - m) / s, df) / s  # Density of the t-distribution with fitted parameters

# Add the t-distribution curve to the plot
lines(x_vals, t_density, col = "red", lwd = 2)

# Overlay the density of the normal distribution
normal_density <- dnorm(x_vals, mean = mean_normal, sd = sd_normal)

# Add the normal distribution curve to the plot
lines(x_vals, normal_density, col = "blue", lwd = 2)

# Add a legend to distinguish between t-distribution and normal distribution
legend("topright", legend = c("Fitted t-Distribution", "Normal Distribution"), 
       col = c("red", "blue"), lwd = 2)
```

#### Figure 8: Density Comparison: Normal vs t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Calculate mean and standard deviation of the negative log returns
mean_neg_log_returns <- mean(bitcoin_negative_log_returns)
sd_neg_log_returns <- sd(bitcoin_negative_log_returns)

# Create a sequence of values for plotting the densities
x_vals <- seq(min(bitcoin_negative_log_returns), max(bitcoin_negative_log_returns), length.out = 1000)

# Calculate density for the normal distribution with the same mean and standard deviation
normal_density <- dnorm(x_vals, mean = mean_neg_log_returns, sd = sd_neg_log_returns)

# Calculate density for the t-distribution with the fitted degrees of freedom
t_density <- dt((x_vals - mean_neg_log_returns) / sd_neg_log_returns, df = df) / sd_neg_log_returns

# Plot the densities to compare tails
plot(x_vals, normal_density, type = "l", col = "blue", lwd = 2, 
     main = "Figure 8: Density Comparison: Normal vs t-Distribution",
     ylab = "Density", xlab = "Negative Log Returns")
lines(x_vals, t_density, col = "red", lwd = 2)
legend("topright", legend = c("Normal", "t-Distribution"), col = c("blue", "red"), lwd = 2)
```

#### Figure 9: ACF of Bitcoin Prices

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the ACF of the raw Bitcoin series
ggAcf(bitcoin_prices, main = "ACF of Raw Bitcoin Prices")
```

#### Figure 10: ACF of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the ACF of the negative log returns
ggAcf(bitcoin_negative_log_returns, main = "ACF of Negative Log Returns")
```

#### Figure 11: PACF of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the PACF of the negative log returns
ggPacf(bitcoin_negative_log_returns, main = "PACF of Negative Log Returns")
```

#### Figure 12: ACF of ARIMA(2, 0, 2) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
arima_fit <- arima(bitcoin_negative_log_returns, order = c(2,0,2))
print(arima_fit)

# Extract the residuals
residuals_arima <- residuals(arima_fit)

# ACF of residuals (to check for autocorrelation)
acf(residuals_arima, main = "ACF of ARIMA Model Residuals")
```

#### Figure 13: QQ-Plot of ARIMA(2, 0, 2) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
# QQ-plot of residuals (to check for normality)
qqnorm(residuals_arima)
qqline(residuals_arima, col = "red")
```

#### Figure 14: Residuals of ARIMA(2, 0, 2) Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot residuals over time (to check for changing variance)
plot(residuals_arima, main = "Residuals of ARIMA Model", ylab = "Residuals", xlab = "Time")
```

#### Figure 15: ACF of GARCH Normal(1, 1)

```{r, fig.align='center', fig.width=7, fig.height=4}
### Fitting a GARCH(1,1) model with a normal distribution
garch_normal_fit <- garchFit(~ garch(1, 1), data = bitcoin_negative_log_returns, cond.dist = "norm", trace = FALSE)

# Extract residuals from the fitted GARCH model with normal distribution
garch_normal_residuals <- residuals(garch_normal_fit)

# Plot ACF of residuals to check for autocorrelation
acf(garch_normal_residuals, main = "ACF of Residuals (GARCH Normal)")
```

#### Figure 16: ACF of GARCH t-Distribution(1, 1)

```{r, fig.align='center', fig.width=7, fig.height=4}
### Fitting a GARCH(1,1) model with a standardized t-distribution ###
garch_t_fit <- garchFit(~ garch(1, 1), data = bitcoin_negative_log_returns, cond.dist = "std", trace = FALSE)

# Extract residuals from the fitted GARCH model with t-distribution
garch_t_residuals <- residuals(garch_t_fit)

# Plot ACF of residuals to check for autocorrelation
acf(garch_t_residuals, main = "ACF of Residuals (GARCH t-Distribution)")
```

#### Figure 17: QQ-Plot of GARCH Normal(1, 1) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
# QQ-Plot to check for normality of residuals
qqnorm(garch_normal_residuals, main = "QQ-Plot of Residuals (GARCH Normal)")
qqline(garch_normal_residuals, col = "red")
```

#### Figure 18: QQ-Plot of GARCH t-Distribution(1, 1) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
# Generate a QQ-plot for the t-distribution residuals
df_t <- garch_t_fit@fit$par["shape"]  # Degrees of freedom for the t-distribution
qqplot(qt(ppoints(length(garch_t_residuals)), df = df_t), 
       garch_t_residuals, main = "QQ-Plot of Residuals (GARCH t-Distribution vs t-Quantiles)",
       xlab = "Theoretical Quantiles (t-distribution)", ylab = "Sample Quantiles")
# t-Distribution GARCH Model follow the t-distribution quite well meaning it captures the heavy tails of the data.

# Add a 45-degree line
qqline(garch_t_residuals, distribution = function(p) qt(p, df = df_t), col = "red")
```

#### Figure 19: Cross-Correlation Function (CCF) between Bitcoin and Ethereum

```{r, fig.align='center', fig.width=7, fig.height=4}
# We take the concerned column.
eth_prices <- Crypto_data$Ethereum

# Function to compute negative log returns:
negative_log_returns <- function(prices) {
  log_returns <- diff(log(prices))  # Calculate log returns
  return(-log_returns)  # Return the negative log returns
}

# Apply the function to the ETH prices
eth_negative_log_returns <- negative_log_returns(eth_prices)

# Ensure the length of both series is the same by trimming if necessary
min_length <- min(length(bitcoin_negative_log_returns), length(eth_negative_log_returns))
bitcoin_negative_log_returns <- bitcoin_negative_log_returns[1:min_length]
eth_negative_log_returns <- eth_negative_log_returns[1:min_length]

# Calculate the Cross-Correlation Function (CCF)
ccf_result <- ccf(bitcoin_negative_log_returns, eth_negative_log_returns, plot=TRUE)
```

## Results tables

#### Table 1: Augmented Dickey-Fuller Test for Bitcoin Prices

```{r}
# Perform the ADF test
adf_test_result <- adf.test(bitcoin_prices)

# Print the ADF test result
print(adf_test_result)
```

#### Table 2: Augmented Dickey-Fuller Test for Negative Log Returns

```{r}
# Perform the ADF test on negative log returns
adf_test_neg_log <- adf.test(bitcoin_negative_log_returns)

# Print the result
print(adf_test_neg_log)
```

#### Table 3: Anderson-Darling Test for Normality of Negative Log Returns

```{r}
# Perform the Anderson-Darling test
ad_test_result <- ad.test(bitcoin_negative_log_returns)

# Print the test result
print(ad_test_result)
```

#### Table 4: Ljung-Box Test for Autocorrelation in Bitcoin Prices

```{r}
# Perform Ljung-Box test for the raw Bitcoin series
ljung_box_raw <- Box.test(bitcoin_prices, lag = 20, type = "Ljung-Box") # lag = 20 because rule of thumb n/10

# Print the results for the raw series
print(ljung_box_raw)
```

#### Table 5: Ljung-Box Test for Autocorrelation in Negative Log Returns

```{r}
# Perform Ljung-Box test for the negative log returns
ljung_box_neg_log <- Box.test(bitcoin_negative_log_returns, lag = 20, type = "Ljung-Box")

# Print the results for the negative log returns
print(ljung_box_neg_log)
```

#### Table 6: Ljung-Box Test for ARIMA(2, 0, 2) Residuals

```{r}
# Ljung-Box test on residuals (to formally test for autocorrelation)
Box.test(residuals_arima, lag = 20, type = "Ljung-Box")
```

#### Table 7: Shapiro-Wilk Test for Normality of ARIMA(2, 0, 2) Residuals

```{r}
# Shapiro-Wilk test (to test for normality)
shapiro.test(residuals_arima)
```

#### Table 8: Ljung-Box Test for GARCH Normal(1, 1) Residuals

```{r}
# Perform Ljung-Box test on residuals
Box.test(garch_normal_residuals, lag = 20, type = "Ljung-Box")
```

#### Table 9: Ljung-Box Test for GARCH t-Distribution(1, 1) Residuals

```{r}
# Perform Ljung-Box test on residuals
Box.test(garch_t_residuals, lag = 20, type = "Ljung-Box")
```

#### Table 10: Shapiro-Wilk Test for Normality of GARCH Normal(1, 1) Residuals

```{r}
shapiro.test(garch_normal_residuals)
```
#### Table 11: Ljung-Box Test for ARIMA-GARCH Residuals

```{r}
# Fit an ARIMA(2,0,2) model on the negative log returns
arima_fit <- arima(bitcoin_negative_log_returns, order = c(2, 0, 2))

# Extract the residuals from the ARIMA model
arima_residuals <- residuals(arima_fit)

# Fit a GARCH(1,1) model on the ARIMA residuals
garch_fit_arima_resid <- garchFit(~ garch(1, 1), data = arima_residuals, trace = FALSE)

garch_residuals <- residuals(garch_fit_arima_resid)

# Box-Ljung test on GARCH residuals
box_ljung_test <- Box.test(garch_residuals, lag = 20, type = "Ljung-Box")
print(box_ljung_test)

```
#### Table 12: Correlation Test for Bitcoin and Ethereum Negative Log Returns

```{r}
# Perform the correlation test between Bitcoin and ETH negative log returns
correlation_test <- cor.test(bitcoin_negative_log_returns, eth_negative_log_returns)

# Print the result of the correlation test
print(correlation_test)
```
#### Table 13: Granger Causality Test: Bitcoin Predicting Ethereum

```{r}
# Choose optimal lag length based on criteria
lag_selection <- VARselect(Crypto_data, lag.max = 20, type = "const")
print(lag_selection) # order s= 6 based on this

# Granger causality test for Bitcoin predicting ETH
grangertest(eth_negative_log_returns ~ bitcoin_negative_log_returns, order = 6)
```

#### Table 14: Granger Causality Test: Ethereum Predicting Bitcoin

```{r}
# Granger causality test for ETH predicting Bitcoin
grangertest(bitcoin_negative_log_returns ~ eth_negative_log_returns, order = 6)
```

