---
title: "Risk Analytics - Practical 1"
subtitle : "Winter semester 2024-2025, HEC, UNIL"
author: "Robin Michel, Faber Bickerstaffe, Antoine Magnin, Anastasia Pushkarev and Victorien Rodondi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Load libraries
library(readr)
library(tseries)
library(nortest)
library(MASS)
library(fpp2)
library(fGarch)
library(lmtest)
library(ggplot2)
library(tidyr)

# Load data using a relative path
Crypto_data <- read_csv("Data/Crypto_data.csv")
```

# Part 1: Financial Returns and Normality

### a) and b) Assessing Stationarity of Bitcoin Prices and Negative Log Returns

In this section, we assess the stationarity of the raw Bitcoin prices and their negative log returns (see Figure 1), as stationarity is crucial for time series modeling.

We first plotted the Bitcoin prices (see Figure 2) to visually inspect for trends or patterns. The plot showed a clear trend, suggesting non-stationarity. To confirm this, we applied the Augmented Dickey-Fuller (ADF) test (see Table 1), which resulted in a p-value of 0.3885, indicating that the raw Bitcoin prices are non-stationary.

To address this, we computed the negative log returns, a transformation commonly used in financial time series analysis to obtain a stationary series. Visual inspection of the negative log returns (see Figure 3) suggested stationarity. This was further confirmed by the ADF test (see Table 2), which gave a p-value of 0.01, leading us to reject the null hypothesis of non-stationarity and confirming that the negative log returns are stationary.

```{r, fig.align='center', fig.width=7, fig.height=2}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Create a data frame for both Bitcoin prices and negative log returns
bitcoin_data <- data.frame(
  Time = 1:length(bitcoin_prices),
  Bitcoin_Prices = bitcoin_prices,
  Negative_Log_Returns = c(NA, bitcoin_negative_log_returns) # Aligning lengths by adding NA
)

# Reshape data for facet wrapping
bitcoin_data_long <- pivot_longer(bitcoin_data, cols = c(Bitcoin_Prices, Negative_Log_Returns), 
                                  names_to = "Series", values_to = "Values")

# Create the facet plot
ggplot(bitcoin_data_long, aes(x = Time, y = Values)) +
  geom_line() +
  facet_wrap(~ Series, scales = "free_y") +
  labs(title = "Figure 2 and 3: Bitcoin Prices and Negative Log Returns Over Time", 
       x = "Time", y = "Value") +
  theme_minimal()
```

### c) Assessing the Normality of Negative Log Returns

To evaluate whether the negative log returns of Bitcoin follow a normal distribution, we first examined the data visually using a histogram (see Figure 4) and a QQ-plot (see Figure 5). The histogram of the negative log returns suggests that the data may be approximately normally distributed, though some deviations from normality could exist. Additionally, the QQ-plot shows that the returns are normally distributed for non-extreme values, but there are noticeable deviations in the tails, indicating that the negative log returns may not follow a perfect normal distribution.

To formally test for normality, we applied the Anderson-Darling test (see Table 3), which gave a p-value less than 0.05. As a result, we reject the null hypothesis (H0) that the data is normally distributed. This confirms that, despite appearing somewhat normal in the central part of the distribution, the negative log returns are not normally distributed, especially due to the extreme values.

### d) and e) Fitting a t-Distribution and Comparing Tails

Since the negative log returns deviate from normality, particularly in the extremes, we fit a t-distribution to the scaled data to check if it better captures these extreme values. A QQ-plot was generated to compare the negative log returns with the theoretical t-distribution (see Figure 6), which showed that the data fits the t-distribution quite well, including in the tails. For comparison, we also generated a QQ-plot for the normal distribution (see Figure 5), which demonstrated a poorer fit, particularly for extreme values. This suggests that the t-distribution, with its ability to model heavy tails, is a more appropriate fit for the data (see Figure 8).

Next, we compared the density plots of the normal and t-distributions. As expected, the t-distribution exhibited heavier tails than the normal distribution, meaning we should expect more extreme, unexpected events in a t-distribution (see Figure 9).

```{r, fig.align='center', fig.width=7, fig.height=3}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Calculate mean and standard deviation of the negative log returns
mean_neg_log_returns <- mean(bitcoin_negative_log_returns)
sd_neg_log_returns <- sd(bitcoin_negative_log_returns)

# Create a sequence of values for plotting the densities
x_vals <- seq(min(bitcoin_negative_log_returns), max(bitcoin_negative_log_returns), length.out = 1000)

# Calculate density for the normal distribution with the same mean and standard deviation
normal_density <- dnorm(x_vals, mean = mean_neg_log_returns, sd = sd_neg_log_returns)

# Calculate density for the t-distribution with the fitted degrees of freedom
t_density <- dt((x_vals - mean_neg_log_returns) / sd_neg_log_returns, df = df) / sd_neg_log_returns

# Plot the densities to compare tails
plot(x_vals, normal_density, type = "l", col = "blue", lwd = 2, 
     main = "Figure 8: Density Comparison: Normal vs t-Distribution",
     ylab = "Density", xlab = "Negative Log Returns")
lines(x_vals, t_density, col = "red", lwd = 2)
legend("topright", legend = c("Normal", "t-Distribution"), col = c("blue", "red"), lwd = 2)
```

Since the Bitcoin data follows the t-distribution more closely, and the t-distribution has fatter tails than the normal distribution, this indicates that extreme values (large deviations from the mean) are more likely in the Bitcoin data than if it were normally distributed.

# Part 2: Financial time series, heteroscedasticity and the random walk hypothesis

## a) and b) Assessing Autocorrelation in Bitcoin Prices

Building on the previous stationarity analysis, we further explored the autocorrelation of the raw Bitcoin prices and the negative log returns. In part (a), we plotted the ACF (Autocorrelation Function) for both series (see Figures 9 and 10). As expected, the raw Bitcoin prices showed strong and persistent autocorrelation, consistent with their non-stationary nature. In contrast, the negative log returns displayed only minor, short-lived autocorrelation, aligning with their stationary behavior.

In part (b), we applied the Ljung-Box test (see Tables 4 and 5) to formally assess autocorrelation. The test confirmed significant autocorrelation in both series (p-value < 0.05). However, the negative log returns exhibited much weaker and short-term autocorrelation compared to the raw prices. This suggests that, while the log returns are stationary, they still contain some mild autocorrelation, typical of financial time series.

In summary, the raw series has strong autocorrelation and is non-stationary, whereas the negative log returns are stationary with only moderate, short-lived autocorrelation, making them better suited for modeling.


## c) ARIMA Model 

In this section, we proposed an ARIMA model for the negative log returns based on the analysis of the ACF and PACF plots (see Figures 10 and 11). The ACF plot showed a spike at lag 2, suggesting a moving average component of q = 2. Similarly, the PACF plot indicated a significant spike at lag 2, suggesting an autoregressive component of p = 2. Since the series is already stationary, we set d = 0. Based on this, we fit an ARIMA(2, 0, 2) model to the negative log returns. To confirm our model selection, we used the auto.arima() function, which automatically identifies the best ARIMA model. The function also suggested an ARIMA(2, 0, 2) model, reinforcing our choice.

Next, we assessed the residuals of the ARIMA model to evaluate its adequacy. The ACF of the residuals showed no significant autocorrelation (see Figure 12), indicating that the model successfully captured the underlying patterns in the data. The Ljung-Box test further confirmed this (see Table 6), as the p-value was greater than 0.05, suggesting no significant autocorrelation remaining in the residuals. However, the QQ-plot (see Figure 13) and Shapiro-Wilk test (see Table 7) indicated that the residuals do not follow a normal distribution (p-value < 0.05). Additionally, the plot of residuals over time (see Figure 14) showed signs of volatility clustering, implying that the variance is not constant. This suggests that while the ARIMA model adequately captures the autocorrelation in the data, it fails to account for the changing variance and heavy tails in the residuals.

In conclusion, the ARIMA(2, 0, 2) model effectively captures the serial dependencies in the negative log returns but may not fully address the volatility present in the data. This indicates that a GARCH model, which accounts for time-varying volatility, could be a better fit for modeling the residuals.

# Appendices

## Practical 1

### Figures

#### Figure 1: Bitcoin Prices and Negative Log Returns Over Time on Common Scale

```{r, fig.align='center', fig.width=7, fig.height=4}
# Standardize both series (subtract mean, divide by standard deviation)
bitcoin_prices_standardized <- scale(bitcoin_prices)
bitcoin_negative_log_returns_standardized <- scale(bitcoin_negative_log_returns)

# Create a data frame combining both series for plotting
bitcoin_data_standardized <- data.frame(
  Time = 1:length(bitcoin_prices_standardized),
  Bitcoin_Prices_Standardized = bitcoin_prices_standardized,
  Negative_Log_Returns_Standardized = c(NA, bitcoin_negative_log_returns_standardized)  # Add NA to align lengths
)

# Reshape the data for ggplot
bitcoin_data_long_standardized <- pivot_longer(bitcoin_data_standardized, 
                                               cols = c(Bitcoin_Prices_Standardized, Negative_Log_Returns_Standardized), 
                                               names_to = "Series", values_to = "Values")

# Plot the standardized series on a common scale, ensuring raw prices are drawn last (on top)
ggplot() +
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Negative_Log_Returns_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Bitcoin_Prices_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  labs(title = "Comparison of Standardized Bitcoin Prices and Negative Log Returns",
       x = "Time", y = "Standardized Values") +
  scale_color_manual(values = c("Bitcoin_Prices_Standardized" = "blue", 
                                "Negative_Log_Returns_Standardized" = "red"),
                     labels = c("Prices", "Log Returns")) +  # Shortened labels
  theme_minimal()
```


#### Figure 2: Bitcoin Prices Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Plot the Bitcoin prices to visually inspect stationarity
plot(bitcoin_prices, type = "l", main = "Bitcoin Prices", xlab = "Time", ylab = "Price")
```

#### Figure 3: Negative Log Returns of Bitcoin Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Plot the negative log returns
plot(bitcoin_negative_log_returns, type = "l", main = "Negative Log Returns of Bitcoin", xlab = "Time", ylab = "Negative Log Returns")
```


#### Figure 4: Histogram of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Draw a histogram of the negative log returns
hist(bitcoin_negative_log_returns, breaks = 30, main = "Histogram of Negative Log Returns", xlab = "Negative Log Returns", col = "lightblue", border = "black")
# --> seems normally distributed
```

#### Figure 5: QQ-Plot of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a QQ-plot to check normality
qqnorm(bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs. Normal Distribution")
qqline(bitcoin_negative_log_returns, col = "red") # --> normally distributed only for non-extreme values
```

#### Figure 6: QQ-Plot of Negative Log Returns with t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
### QQ-plot for t-distribution : seem to follow it quite well
# Generate a QQ-plot for the t-distribution fit
qqplot(qt(ppoints(length(bitcoin_negative_log_returns)), df = df_value), 
       bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs t-Distribution",
       xlab = "Theoretical Quantiles (t-distribution)", ylab = "Sample Quantiles")
# Add a 45-degree line to the QQ-plot
qqline(bitcoin_negative_log_returns, distribution = function(p) qt(p, df = df_value), col = "red")
```

#### Figure 7: Histogram of Bitcoin Negative Log Returns with Fitted t and Normal Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract the fitted parameters from t_fit
m <- t_fit$estimate["m"]      # Mean (location parameter)
s <- t_fit$estimate["s"]      # Scale (related to standard deviation)
df <- t_fit$estimate["df"]    # Degrees of freedom

# Calculate the mean and standard deviation for the normal distribution
mean_normal <- mean(bitcoin_negative_log_returns_scaled)
sd_normal <- sd(bitcoin_negative_log_returns_scaled)

# Plot the histogram of your data
hist(bitcoin_negative_log_returns_scaled, breaks = 30, probability = TRUE, 
     main = "Histogram of Bitcoin Neg. Log Ret. with Fitted t and Normal Distribution",
     xlab = "Negative Log Returns", col = "lightblue", border = "black")

# Overlay the density of the fitted t-distribution
x_vals <- seq(min(bitcoin_negative_log_returns_scaled), max(bitcoin_negative_log_returns_scaled), length.out = 1000)
t_density <- dt((x_vals - m) / s, df) / s  # Density of the t-distribution with fitted parameters

# Add the t-distribution curve to the plot
lines(x_vals, t_density, col = "red", lwd = 2)

# Overlay the density of the normal distribution
normal_density <- dnorm(x_vals, mean = mean_normal, sd = sd_normal)

# Add the normal distribution curve to the plot
lines(x_vals, normal_density, col = "blue", lwd = 2)

# Add a legend to distinguish between t-distribution and normal distribution
legend("topright", legend = c("Fitted t-Distribution", "Normal Distribution"), 
       col = c("red", "blue"), lwd = 2)
```

#### Figure 8: Density Comparison: Normal vs t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Calculate mean and standard deviation of the negative log returns
mean_neg_log_returns <- mean(bitcoin_negative_log_returns)
sd_neg_log_returns <- sd(bitcoin_negative_log_returns)

# Create a sequence of values for plotting the densities
x_vals <- seq(min(bitcoin_negative_log_returns), max(bitcoin_negative_log_returns), length.out = 1000)

# Calculate density for the normal distribution with the same mean and standard deviation
normal_density <- dnorm(x_vals, mean = mean_neg_log_returns, sd = sd_neg_log_returns)

# Calculate density for the t-distribution with the fitted degrees of freedom
t_density <- dt((x_vals - mean_neg_log_returns) / sd_neg_log_returns, df = df) / sd_neg_log_returns

# Plot the densities to compare tails
plot(x_vals, normal_density, type = "l", col = "blue", lwd = 2, 
     main = "Figure 8: Density Comparison: Normal vs t-Distribution",
     ylab = "Density", xlab = "Negative Log Returns")
lines(x_vals, t_density, col = "red", lwd = 2)
legend("topright", legend = c("Normal", "t-Distribution"), col = c("blue", "red"), lwd = 2)
```

#### Figure 9: ACF of Bitcoin Prices

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the ACF of the raw Bitcoin series
ggAcf(bitcoin_prices, main = "ACF of Raw Bitcoin Prices")
```

#### Figure 10: ACF of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the ACF of the negative log returns
ggAcf(bitcoin_negative_log_returns, main = "ACF of Negative Log Returns")
```

#### Figure 11: PACF of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot the PACF of the negative log returns
ggPacf(bitcoin_negative_log_returns, main = "PACF of Negative Log Returns")
```

#### Figure 12: ACF of ARIMA(2, 0, 2) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
arima_fit <- arima(bitcoin_negative_log_returns, order = c(2,0,2))
print(arima_fit)

# Extract the residuals
residuals_arima <- residuals(arima_fit)

# ACF of residuals (to check for autocorrelation)
acf(residuals_arima, main = "ACF of ARIMA Model Residuals")
```

#### Figure 13: QQ-Plot of ARIMA(2, 0, 2) Residuals

```{r, fig.align='center', fig.width=7, fig.height=4}
# QQ-plot of residuals (to check for normality)
qqnorm(residuals_arima)
qqline(residuals_arima, col = "red")
```

#### Figure 14: Residuals of ARIMA(2, 0, 2) Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Plot residuals over time (to check for changing variance)
plot(residuals_arima, main = "Residuals of ARIMA Model", ylab = "Residuals", xlab = "Time")
```

## Results tables

#### Table 1: Augmented Dickey-Fuller Test for Bitcoin Prices

```{r}
# Perform the ADF test
adf_test_result <- adf.test(bitcoin_prices)

# Print the ADF test result
print(adf_test_result)
```

#### Table 2: Augmented Dickey-Fuller Test for Negative Log Returns

```{r}
# Perform the ADF test on negative log returns
adf_test_neg_log <- adf.test(bitcoin_negative_log_returns)

# Print the result
print(adf_test_neg_log)
```

#### Table 3: Anderson-Darling Test for Normality of Negative Log Returns

```{r}
# Perform the Anderson-Darling test
ad_test_result <- ad.test(bitcoin_negative_log_returns)

# Print the test result
print(ad_test_result)
```

#### Table 4: Ljung-Box Test for Autocorrelation in Bitcoin Prices

```{r}
# Perform Ljung-Box test for the raw Bitcoin series
ljung_box_raw <- Box.test(bitcoin_prices, lag = 20, type = "Ljung-Box") # lag = 20 because rule of thumb n/10

# Print the results for the raw series
print(ljung_box_raw)
```

#### Table 5: Ljung-Box Test for Autocorrelation in Negative Log Returns

```{r}
# Perform Ljung-Box test for the negative log returns
ljung_box_neg_log <- Box.test(bitcoin_negative_log_returns, lag = 20, type = "Ljung-Box")

# Print the results for the negative log returns
print(ljung_box_neg_log)
```

#### Table 6: Ljung-Box Test for ARIMA(2, 0, 2) Residuals

```{r}
# Ljung-Box test on residuals (to formally test for autocorrelation)
Box.test(residuals_arima, lag = 20, type = "Ljung-Box")
```

#### Table 7: Shapiro-Wilk Test for Normality of ARIMA(2, 0, 2) Residuals

```{r}
# Shapiro-Wilk test (to test for normality)
shapiro.test(residuals_arima)
```

# Code Appendix
