---
title: "Risk Analytics - Practical 1"
subtitle : "Winter semester 2024-2025, HEC, UNIL"
author: "Robin Michel, Faber Bickerstaffe, Antoine Magnin, Anastasia Pushkarev and Victorien Rodondi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Load libraries
library(readr)
library(tseries)
library(nortest)
library(MASS)
library(fpp2)
library(fGarch)
library(lmtest)
library(ggplot2)
library(tidyr)

# Load data using a relative path
Crypto_data <- read_csv("Data/Crypto_data.csv")
```

# Part 1: Financial Returns and Normality

### a) and b) Assessing Stationarity of Bitcoin Prices and Negative Log Returns

In this section, we assess the stationarity of the raw Bitcoin prices and their negative log returns (see Figure 1), as stationarity is crucial for time series modeling.

We first plotted the Bitcoin prices (see Figure 2) to visually inspect for trends or patterns. The plot showed a clear trend, suggesting non-stationarity. To confirm this, we applied the Augmented Dickey-Fuller (ADF) test (see Table 1), which resulted in a p-value of 0.3885, indicating that the raw Bitcoin prices are non-stationary.

To address this, we computed the negative log returns, a transformation commonly used in financial time series analysis to obtain a stationary series. Visual inspection of the negative log returns (see Figure 3) suggested stationarity. This was further confirmed by the ADF test (see Table 2), which gave a p-value of 0.01, leading us to reject the null hypothesis of non-stationarity and confirming that the negative log returns are stationary.

```{r, fig.align='center', fig.width=7, fig.height=2}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Create a data frame for both Bitcoin prices and negative log returns
bitcoin_data <- data.frame(
  Time = 1:length(bitcoin_prices),
  Bitcoin_Prices = bitcoin_prices,
  Negative_Log_Returns = c(NA, bitcoin_negative_log_returns) # Aligning lengths by adding NA
)

# Reshape data for facet wrapping
bitcoin_data_long <- pivot_longer(bitcoin_data, cols = c(Bitcoin_Prices, Negative_Log_Returns), 
                                  names_to = "Series", values_to = "Values")

# Create the facet plot
ggplot(bitcoin_data_long, aes(x = Time, y = Values)) +
  geom_line() +
  facet_wrap(~ Series, scales = "free_y") +
  labs(title = "Figure 2 and 3: Bitcoin Prices and Negative Log Returns Over Time", 
       x = "Time", y = "Value") +
  theme_minimal()
```

### c) Assessing the Normality of Negative Log Returns

To evaluate whether the negative log returns of Bitcoin follow a normal distribution, we first examined the data visually using a histogram (see Figure 4) and a QQ-plot (see Figure 5). The histogram of the negative log returns suggests that the data may be approximately normally distributed, though some deviations from normality could exist. Additionally, the QQ-plot shows that the returns are normally distributed for non-extreme values, but there are noticeable deviations in the tails, indicating that the negative log returns may not follow a perfect normal distribution.

To formally test for normality, we applied the Anderson-Darling test (see Table 3), which gave a p-value less than 0.05. As a result, we reject the null hypothesis (H0) that the data is normally distributed. This confirms that, despite appearing somewhat normal in the central part of the distribution, the negative log returns are not normally distributed, especially due to the extreme values.

### d) and e) Fitting a t-Distribution and Comparing Tails

Since the negative log returns deviate from normality, particularly in the extremes, we fit a t-distribution to the scaled data to check if it better captures these extreme values. A QQ-plot was generated to compare the negative log returns with the theoretical t-distribution (see Figure 6), which showed that the data fits the t-distribution quite well, including in the tails. For comparison, we also generated a QQ-plot for the normal distribution (see Figure 5), which demonstrated a poorer fit, particularly for extreme values. This suggests that the t-distribution, with its ability to model heavy tails, is a more appropriate fit for the data (see Figure 8).

Next, we compared the density plots of the normal and t-distributions. As expected, the t-distribution exhibited heavier tails than the normal distribution, meaning we should expect more extreme, unexpected events in a t-distribution (see Figure 9).

```{r, fig.align='center', fig.width=7, fig.height=3}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Calculate mean and standard deviation of the negative log returns
mean_neg_log_returns <- mean(bitcoin_negative_log_returns)
sd_neg_log_returns <- sd(bitcoin_negative_log_returns)

# Create a sequence of values for plotting the densities
x_vals <- seq(min(bitcoin_negative_log_returns), max(bitcoin_negative_log_returns), length.out = 1000)

# Calculate density for the normal distribution with the same mean and standard deviation
normal_density <- dnorm(x_vals, mean = mean_neg_log_returns, sd = sd_neg_log_returns)

# Calculate density for the t-distribution with the fitted degrees of freedom
t_density <- dt((x_vals - mean_neg_log_returns) / sd_neg_log_returns, df = df) / sd_neg_log_returns

# Plot the densities to compare tails
plot(x_vals, normal_density, type = "l", col = "blue", lwd = 2, 
     main = "Figure 8: Density Comparison: Normal vs t-Distribution",
     ylab = "Density", xlab = "Negative Log Returns")
lines(x_vals, t_density, col = "red", lwd = 2)
legend("topright", legend = c("Normal", "t-Distribution"), col = c("blue", "red"), lwd = 2)
```

Since the Bitcoin data follows the t-distribution more closely, and the t-distribution has fatter tails than the normal distribution, this indicates that extreme values (large deviations from the mean) are more likely in the Bitcoin data than if it were normally distributed.

# Part 2: Financial time series, heteroscedasticity and the random walk hypothesis

# Appendices

## Practical 1

### Figures

#### Figure 1: Bitcoin Prices and Negative Log Returns Over Time on Common Scale

```{r, fig.align='center', fig.width=7, fig.height=4}
# Standardize both series (subtract mean, divide by standard deviation)
bitcoin_prices_standardized <- scale(bitcoin_prices)
bitcoin_negative_log_returns_standardized <- scale(bitcoin_negative_log_returns)

# Create a data frame combining both series for plotting
bitcoin_data_standardized <- data.frame(
  Time = 1:length(bitcoin_prices_standardized),
  Bitcoin_Prices_Standardized = bitcoin_prices_standardized,
  Negative_Log_Returns_Standardized = c(NA, bitcoin_negative_log_returns_standardized)  # Add NA to align lengths
)

# Reshape the data for ggplot
bitcoin_data_long_standardized <- pivot_longer(bitcoin_data_standardized, 
                                               cols = c(Bitcoin_Prices_Standardized, Negative_Log_Returns_Standardized), 
                                               names_to = "Series", values_to = "Values")

# Plot the standardized series on a common scale, ensuring raw prices are drawn last (on top)
ggplot() +
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Negative_Log_Returns_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  geom_line(data = subset(bitcoin_data_long_standardized, Series == "Bitcoin_Prices_Standardized"), 
            aes(x = Time, y = Values, color = Series), size = 0.5) +  
  labs(title = "Comparison of Standardized Bitcoin Prices and Negative Log Returns",
       x = "Time", y = "Standardized Values") +
  scale_color_manual(values = c("Bitcoin_Prices_Standardized" = "blue", 
                                "Negative_Log_Returns_Standardized" = "red"),
                     labels = c("Prices", "Log Returns")) +  # Shortened labels
  theme_minimal()
```


#### Figure 2: Bitcoin Prices Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract Bitcoin prices 
bitcoin_prices <- Crypto_data[[1]] # or Crypto_data$Bitcoin

# Plot the Bitcoin prices to visually inspect stationarity
plot(bitcoin_prices, type = "l", main = "Bitcoin Prices", xlab = "Time", ylab = "Price")
```

#### Figure 3: Negative Log Returns of Bitcoin Over Time

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a function to compute negative log returns
negative_log_returns <- function(prices) {
  # Calculate log returns
  log_returns <- diff(log(prices))
  # Return the negative of the log returns
  return(-log_returns)
}

# Apply the function to the Bitcoin prices
bitcoin_negative_log_returns <- negative_log_returns(bitcoin_prices)

# Plot the negative log returns
plot(bitcoin_negative_log_returns, type = "l", main = "Negative Log Returns of Bitcoin", xlab = "Time", ylab = "Negative Log Returns")
```


#### Figure 4: Histogram of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Draw a histogram of the negative log returns
hist(bitcoin_negative_log_returns, breaks = 30, main = "Histogram of Negative Log Returns", xlab = "Negative Log Returns", col = "lightblue", border = "black")
# --> seems normally distributed
```

#### Figure 5: QQ-Plot of Negative Log Returns

```{r, fig.align='center', fig.width=7, fig.height=4}
# Create a QQ-plot to check normality
qqnorm(bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs. Normal Distribution")
qqline(bitcoin_negative_log_returns, col = "red") # --> normally distributed only for non-extreme values
```

#### Figure 6: QQ-Plot of Negative Log Returns with t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
### QQ-plot for t-distribution : seem to follow it quite well
# Generate a QQ-plot for the t-distribution fit
qqplot(qt(ppoints(length(bitcoin_negative_log_returns)), df = df_value), 
       bitcoin_negative_log_returns, main = "QQ-Plot of Negative Log Returns vs t-Distribution",
       xlab = "Theoretical Quantiles (t-distribution)", ylab = "Sample Quantiles")
# Add a 45-degree line to the QQ-plot
qqline(bitcoin_negative_log_returns, distribution = function(p) qt(p, df = df_value), col = "red")
```

#### Figure 7: Histogram of Bitcoin Negative Log Returns with Fitted t and Normal Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Extract the fitted parameters from t_fit
m <- t_fit$estimate["m"]      # Mean (location parameter)
s <- t_fit$estimate["s"]      # Scale (related to standard deviation)
df <- t_fit$estimate["df"]    # Degrees of freedom

# Calculate the mean and standard deviation for the normal distribution
mean_normal <- mean(bitcoin_negative_log_returns_scaled)
sd_normal <- sd(bitcoin_negative_log_returns_scaled)

# Plot the histogram of your data
hist(bitcoin_negative_log_returns_scaled, breaks = 30, probability = TRUE, 
     main = "Histogram of Bitcoin Neg. Log Ret. with Fitted t and Normal Distribution",
     xlab = "Negative Log Returns", col = "lightblue", border = "black")

# Overlay the density of the fitted t-distribution
x_vals <- seq(min(bitcoin_negative_log_returns_scaled), max(bitcoin_negative_log_returns_scaled), length.out = 1000)
t_density <- dt((x_vals - m) / s, df) / s  # Density of the t-distribution with fitted parameters

# Add the t-distribution curve to the plot
lines(x_vals, t_density, col = "red", lwd = 2)

# Overlay the density of the normal distribution
normal_density <- dnorm(x_vals, mean = mean_normal, sd = sd_normal)

# Add the normal distribution curve to the plot
lines(x_vals, normal_density, col = "blue", lwd = 2)

# Add a legend to distinguish between t-distribution and normal distribution
legend("topright", legend = c("Fitted t-Distribution", "Normal Distribution"), 
       col = c("red", "blue"), lwd = 2)
```


#### Figure 8: Density Comparison: Normal vs t-Distribution

```{r, fig.align='center', fig.width=7, fig.height=4}
# Scale the negative log returns
bitcoin_negative_log_returns_scaled <- scale(bitcoin_negative_log_returns)

# Fit a t-distribution to the negative log returns
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")

# Extract the degrees of freedom (df) from the fitted distribution
df_value <- t_fit$estimate["df"]

# Recalculate degrees of freedom within the same chunk
t_fit <- fitdistr(bitcoin_negative_log_returns_scaled, "t")
df <- t_fit$estimate["df"]

# Calculate mean and standard deviation of the negative log returns
mean_neg_log_returns <- mean(bitcoin_negative_log_returns)
sd_neg_log_returns <- sd(bitcoin_negative_log_returns)

# Create a sequence of values for plotting the densities
x_vals <- seq(min(bitcoin_negative_log_returns), max(bitcoin_negative_log_returns), length.out = 1000)

# Calculate density for the normal distribution with the same mean and standard deviation
normal_density <- dnorm(x_vals, mean = mean_neg_log_returns, sd = sd_neg_log_returns)

# Calculate density for the t-distribution with the fitted degrees of freedom
t_density <- dt((x_vals - mean_neg_log_returns) / sd_neg_log_returns, df = df) / sd_neg_log_returns

# Plot the densities to compare tails
plot(x_vals, normal_density, type = "l", col = "blue", lwd = 2, 
     main = "Figure 8: Density Comparison: Normal vs t-Distribution",
     ylab = "Density", xlab = "Negative Log Returns")
lines(x_vals, t_density, col = "red", lwd = 2)
legend("topright", legend = c("Normal", "t-Distribution"), col = c("blue", "red"), lwd = 2)
```



## Results tables

#### Table 1: Augmented Dickey-Fuller Test for Bitcoin Prices

```{r}
# Perform the ADF test
adf_test_result <- adf.test(bitcoin_prices)

# Print the ADF test result
print(adf_test_result)
```

#### Table 2: Augmented Dickey-Fuller Test for Negative Log Returns

```{r}
# Perform the ADF test on negative log returns
adf_test_neg_log <- adf.test(bitcoin_negative_log_returns)

# Print the result
print(adf_test_neg_log)
```

#### Table 3: Anderson-Darling Test for Normality of Negative Log Returns

```{r}
# Perform the Anderson-Darling test
ad_test_result <- ad.test(bitcoin_negative_log_returns)

# Print the test result
print(ad_test_result)
```

# Code Appendix
